Time-series modeling of epidemics in complex populations: detecting changes in incidence volatility over time

### 2024-12-19 ms-3/ms.tex
Complete intro

\begin{abstract}
Trends in infectious disease incidence provide important information about epidemic dynamics and prospects for control. 
Higher-frequency variation around incidence trends can shed light on the processes driving epidemics in complex populations, as transmission heterogeneity, shifting landscapes of susceptibility, and fluctuations in reporting can impact the volatility of observed case counts.
However, measures of temporal volatility in incidence, and how it changes over time, is often overlooked in population-level analyses of incidence data, which typically focus on moving averages.
Here we present a statistical framework to quantify temporal changes in incidence dispersion and detect discrete shifts in the dispersion parameter, which may signal new epidemic phases. 
We apply the method to COVID-19 incidence data in 144 US counties from the January 1st, 2020 to March 23rd, 2023.
Theory predicts that dispersion should be inversely proportional to incidence, however our method reveals pronounced temporal trends in dispersion that are not explained by incidence alone, but which are replicated across counties. 
In particular, dispersion increased around the major surge in cases in 2022, and highly overdispersed patterns became more frequent later in the time series.
These findings suggest that heterogeneity in transmission, susceptibility, and reporting could play important roles in driving large surges and extending epidemic duration. 
The dispersion of incidence time series can contain structured information which enhances predictive understanding of the underlying drivers of transmission, with potential applications as leading indicators for public health response.
\end{abstract}

@@

>>>>

\begin{abstract}
Trends in infectious disease incidence provide important information about epidemic dynamics and prospects for control. 
Higher-frequency variation around incidence trends can shed light on the processes driving epidemics in complex populations, as transmission heterogeneity, shifting landscapes of susceptibility, and fluctuations in reporting can impact the volatility of observed case counts.
However, measures of temporal volatility in incidence, and how volatility changes over time, are often overlooked in population-level analyses of incidence data, which typically focus on moving averages.
Here we present a statistical framework to quantify temporal changes in incidence dispersion and detect discrete shifts in the dispersion parameter, which may signal new epidemic phases. 
We apply the method to COVID-19 incidence data in 144 US counties from the January 1st, 2020 to March 23rd, 2023.
Theory predicts that dispersion should be inversely proportional to incidence, however our method reveals pronounced temporal trends in dispersion that are not explained by incidence alone, but which are replicated across counties. 
In particular, dispersion increased around the major surge in cases in 2022, and highly overdispersed patterns became more frequent later in the time series.
These findings suggest that heterogeneity in transmission, susceptibility, and reporting could play important roles in driving large surges and extending epidemic duration. 
The dispersion of incidence time series can contain structured information which enhances predictive understanding of the underlying drivers of transmission, with potential applications as leading indicators for public health response.
\end{abstract}

@@

Emerging statistical techniques are leveraging variability in epidemic time series to enhance understanding of disease dynamics at the population level. 
For example, a recently-developed method estimates the dispersion parameter of the offspring distribution, which quantifies heterogeneity in secondary cases generated by an infected individual, directly from population-level incidence data~\cite{kirkegaard2021superspreading}. 
Clustering of cases has also been estimated directly from incidence data~\cite{schneckenreither2023assessing}.
Another important application links variability in incidence to epidemic phases; for example, changes in the mean and interannual coefficient of variation of measles incidence have been used to identify a country’s position on the path to elimination, providing insights into vaccination strategies and epidemiological dynamics~\cite{graham2019measles}. 
Analysis of the shape of epidemic curves for influenza in cities also aids public health planning by identifying contexts where incidence is focussed more intensely (proportionally more infections in a smaller span of time) with relevance for surge capacity in the health system~\cite{dalziel2018urbanization, wallinga2018metropolitan}.

What drives indicence dispersion and how does it relate to the underlying branching process of transmission, and to observations of cases? 
Under a wide range of configurations for a branching process model the number of infected individuals $I_t$ at time $t$ will have a negative binomial distribution~\cite{kendall1949stochastic, grenfell2002dynamics}, $I_t \sim NB \left( \mu_t, \theta_t \right)$, where $\mu_t$ is the expectation for $I_t$ and $\theta_t$ is the dispersion parameter. 
The variance is related to the mean and dispersion parameters by $\mathrm{Var}[I_t] = \mu_t + \mu_t^2 / \theta_t$, so smaller values of the dispersion parameter $\theta_t$ correspond to increasing amounts of dispersion, which increase the amounts by which the variance in realized number infected $I_t$ exceeds the expected value, $\mu_t$. 
Conversely, the distribution of $I_t$ tends to a Poisson distribution (where the variance equals mean) as $\theta_t$ becomes large. 
The negative binomial distribution may also accurately model a time series if there is a changing process mean within a time step: for example, if the mean of a Poisson distribution itself follows a gamma distribution, the resulting distribution is negative binomial. 
Negative binomial regression (in contrast to Poisson regression) can account for unobserved heterogeneity, time dependence in the rate of a process and contagion within a time step that all lead to overdispersion~\cite{barron1992analysis}.

An interpretation of the dispersion parameter for a time series model of counts is that events are $1 + \theta^{-1}$ times as ``crowded'' in time relative to a Poisson process with the same mean~\cite{lloyd1967mean} (see Supplemental Information). 
For example, $\theta = 1$ corresponds to a situation where the average number of infections in the same time step as a randomly selected case will exceed the Poisson expectation by a factor of two. 
In a simple example relevant to surge capacity in the healthcare system, $\theta = 1$ implies that a random infectious individual visiting the emergency department at a hospital would find it on average to be twice as crowded with other infectious individuals (infected by the same pathogen) than expected for a Poisson process with the same incidence rate.

In a sufficiently large host population, and when the infectious pathogen can be assumed to spread in nonoverlapping generations, incidence is often modeled as
\begin{equation}
    \mathrm{I_{t+1}} = NB(\mu = \lambda R_t I_t, \theta_t = I_t)
\end{equation}
where time-varying reproductive number $R_t$ gives the expected number of secondary infections aqcuired from an infected indivual at time $t$, and the generation time is set to 1 without loss of generality~\cite{kendall1949stochastic, bjornstad2002dynamics}. 
Setting $\theta_t = I_t$ arises from the assumption that individuals who acquire the infection at time $t$ form independent lineages with identically distributed local rate parameters.
However, this requires that susceptible depletion in one lineage does not affect another, that transmission rates are equal across lineages, and that reporting rates do not vary across lineages.

In practice, these assumption will not often hold, and our aim in this paper is to develop, test and apply an alternative approach, which makes data-driven estimates of $\theta_t$, including identifying probably timepoints when $\theta$ is changing rapidly, which may help to reveal how heterogeneity in transmission, susceptibility and reporting, are impacting incidence volatlity.

\section*{Methods}
